---
title: "Magpie Jay Clines - run on KUHPCC"
author: "Devon DeRaad"
date: '2022-10-10'
output: html_document
---

```{r}
## Load the package
library(hzar)

#manually generated set of distances
#c(0,197,447,631,666,1029,1124,1550,1957,1994)

##Load the data for each sample in our historical transect
hyb_Index <- read.csv("Magpie.Jay.data.updated.csv")
#assign population as a factor in a separate column
hyb_Index$sample.loc<-as.factor(hyb_Index$Population)
##sort by distance
#hyb_Index<-hyb_Index[order(hyb_Index$Population),]

#split the dataframe into separate modern and ancestral dataframes based on the 'Period' column
#modern
hyb_Index.modern <- hyb_Index[hyb_Index$Period == 1,]
#ancestral
hyb_Index.anc <- hyb_Index[hyb_Index$Period == 0,]

#make a separate dataframe with geographic distance, and the mean of each input value for each sampling locality
#modern samples only
locs.modern<-data.frame(site.id=unique(hyb_Index.modern$Population),
                 dist=c(0,197,447,631,666,1029,1124,1550,1957,1994),
                 throat=aggregate(hyb_Index.modern$Throat~hyb_Index.modern$Population, FUN=mean)[,2])

#normalize the throat score to a 0-1 range (required by input function)
locs.modern$throat<-locs.modern$throat/5

#make a separate dataframe with geographic distance, and the mean of each input value for each sampling locality
#ancestral samples only
locs.ancestral<-data.frame(site.id=unique(hyb_Index.anc$Population),
                 dist=c(0,197,447,631,666,1029,1124,1550,1957,1994),
                 throat=aggregate(hyb_Index.anc$Throat~hyb_Index.anc$Population, FUN=mean)[,2])

#normalize the throat score to a 0-1 range (required by input function)
locs.ancestral$throat<-locs.ancestral$throat/5

# set Chain length
chainLength=1e6                    
```

### this chunk defines 3 helper functions that will (1) run 3 separate hzar models, (2) check for convergence, and (3) process the cline output to make it easily plot-able
```{r}
### function 1:
#write function to run 3 different hzar models and store the output in a single list
run3hzarmodels<-function(input.trait=NULL, begin=NULL,end=NULL){
  ## create empty object to hold results
  x <- list() #designate the firs trait 'Comp.1'
  x$models <- list() #Space to hold the models to fit
  x$fitRs <- list() #Space to hold the compiled fit requests
  x$runs <- list() #Space to hold the output data chains
  x$analysis <- list() #Space to hold the analysed data
  
  #add input observed data to list
  x$obs<-input.trait
  
  #load the three different models we will test
  #min and max values fixed to observed data, no exponential tails
  x$models[["modelI"]]<-hzar.makeCline1DFreq(x$obs, "fixed", "none")
  #min and max values estimated as free parameters, no exponential tails
  x$models[["modelII"]]<-hzar.makeCline1DFreq(x$obs, "free", "none")
  #min and max values estimated as free paramaters, tails estimated as independent paramaters
  x$models[["modelIII"]]<-hzar.makeCline1DFreq(x$obs, "free", "both")

  #modify all models to focus on observed region 
  x$models <- sapply(x$models, hzar.model.addBoxReq, begin, end, simplify=FALSE)
  
  ## Compile each of the models to prepare for fitting
  #fit each of the 3 models we set up to the observed data
  x$fitRs$init <- sapply(x$models, hzar.first.fitRequest.old.ML, obsData=x$obs, verbose=FALSE, simplify=FALSE)
  
  #update settings for the fitter using chainLength created before
  x$fitRs$init$modelI$mcmcParam$chainLength <- chainLength
  x$fitRs$init$modelI$mcmcParam$burnin <- chainLength %/% 10
  x$fitRs$init$modelII$mcmcParam$chainLength <- chainLength
  x$fitRs$init$modelII$mcmcParam$burnin <- chainLength %/% 10
  x$fitRs$init$modelIII$mcmcParam$chainLength <- chainLength
  x$fitRs$init$modelIII$mcmcParam$burnin <- chainLength %/% 10

  ## Run just one of the models for an initial chain
  x$runs$init$modelI <-hzar.doFit(x$fitRs$init$modelI)
  ## Run another model for an initial chain
  x$runs$init$modelII <- hzar.doFit(x$fitRs$init$modelII)
  ## Run another model for an initial chain
  x$runs$init$modelIII <- hzar.doFit(x$fitRs$init$modelIII)

  ## Compile a new set of fit requests using the initial chains 
  x$fitRs$chains <-lapply(x$runs$init,hzar.next.fitRequest)
  
  ## Replicate each fit request 3 times
  x$fitRs$chains <-hzar.multiFitRequest(x$fitRs$chains,each=3)

  ##Run a chain of 3 runs for every fit request
  x$runs$chains <-hzar.doChain.multi(x$fitRs$chains,doPar=TRUE,inOrder=FALSE,count=3)
  
  return(x)
}

### function 2:
#function to check MCMC convergence
check.convergence<-function(input.hzar=NULL){
  ## Check for convergence
  print("did chains from modelI converge?")
  plot(hzar.mcmc.bindLL(input.hzar$runs$init$modelIII))  ## Plot the trace model I
  print("did chains from modelII converge?")
  plot(hzar.mcmc.bindLL(input.hzar$runs$init$modelIII))  ## Plot the trace model II
  print("did chains from modelIII converge?")
  plot(hzar.mcmc.bindLL(input.hzar$runs$init$modelIII))  ## Plot the trace model III
}

### function 3:
#write function to do the processing necessary before plotting the resulting cline
analyze.hzar.output<-function(input.hzar=NULL, input.var=NULL){
  #add a null model where allele frequency is independent of sampling locality
  input.hzar$analysis$initDGs <- list(nullModel =  hzar.dataGroup.null(input.hzar$obs))

  #start aggregation of data for analysis
  #create a model data group for each model from the initial runs
  input.hzar$analysis$initDGs$modelI<- hzar.dataGroup.add(input.hzar$runs$init$modelI)
  input.hzar$analysis$initDGs$modelII <-hzar.dataGroup.add(input.hzar$runs$init$modelII)
  input.hzar$analysis$initDGs$modelIII<- hzar.dataGroup.add(input.hzar$runs$init$modelIII)
  
  ##create a hzar.obsDataGroup object from the four hzar.dataGroup just created, copying the naming scheme
  input.hzar$analysis$oDG<-hzar.make.obsDataGroup(input.hzar$analysis$initDGs)
  input.hzar$analysis$oDG <- hzar.copyModelLabels(input.hzar$analysis$initDGs,input.hzar$analysis$oDG)
  
  ##convert all runs to hzar.dataGroup objects
  input.hzar$analysis$oDG <-hzar.make.obsDataGroup(input.hzar$analysis$initDGs)
  input.hzar$analysis$oDG <-hzar.copyModelLabels(input.hzar$analysis$initDGs,input.hzar$analysis$oDG)
  input.hzar$analysis$oDG <-hzar.make.obsDataGroup(lapply(input.hzar$runs$chains,hzar.dataGroup.add),input.hzar$analysis$oDG)
  #this no longer works
  #input.hzar$analysis$oDG <- hzar.make.obsDataGroup(lapply(input.hzar$runs$doSeq,   hzar.dataGroup.add),input.hzar$analysis$oDG)
  
  #compare the 5 cline models graphically
  print("output clines from each model overlaid")
  hzar.plot.cline(input.hzar$analysis$oDG)
  
  #model selection based on AICc scores
  print("AICc table")
  print(input.hzar$analysis$AICcTable<- hzar.AICc.hzar.obsDataGroup(input.hzar$analysis$oDG))
  
  #Extract the hzar.dataGroup object for the selected model
  print("best model based on AICc")
  print(input.hzar$analysis$model.name<-   rownames(input.hzar$analysis$AICcTable)[[which.min(input.hzar$analysis$AICcTable$AICc)]])
  input.hzar$analysis$model.selected<- input.hzar$analysis$oDG$data.groups[[input.hzar$analysis$model.name]]
  
  #print the point estimates for cline width and center for the selected model
  input.hzar$analysis$modeldetails <- hzar.get.ML.cline(input.hzar$analysis$model.selected)
  input.hzar$analysis$modeldetails$param.all$width
  input.hzar$analysis$modeldetails$param.all$center
  
  #Print the 2LL confidence intervals for each parameter under the best model
  print("2LL confidence intervals for all estimated parameters")
  print(hzar.getLLCutParam(input.hzar$analysis$model.selected,   names(input.hzar$analysis$model.selected$data.param)))
  
  #plot the maximum likelihood cline for the selected model
  print("maximum likelihood cline")
  hzar.plot.cline(input.hzar$analysis$model.selected,pch=19,xlab="Distance (km)",ylab=input.var)
  
  #plot the 95% credible cline region for the selected model
  print("95% credible cline region for the optimal model")
  hzar.plot.fzCline(input.hzar$analysis$model.selected,pch=19,xlab="Distance (km)",ylab=input.var)
  return(input.hzar)
}
```

### Now we will use these helper functions to make clines for historical throat color
```{r, results='hide'}
## Set up first input trait (throat historical)
throat.hist.input <- hzar.doMolecularData1DPops(distance=locs.ancestral$dist,
                                             pObs=locs.ancestral$throat,
                                             nEff=as.vector(table(hyb_Index.anc$Population)))

#run 3 models
throat.hist<-run3hzarmodels(input.trait=throat.hist.input,
                                 begin=0,
                                 end=1994)
```


```{r, fig.height=10, fig.width=6, fig.align = "center"}
#check convergence
check.convergence(throat.hist)
```

```{r}
#run analysis function on genome wide ancestry input
hist.plot<-analyze.hzar.output(throat.hist, input.var = "historical throat color")
```

### contemporary throat color
```{r, results='hide'}
#Set up next input trait (modern throat color)
throat.contemp.input <- hzar.doMolecularData1DPops(distance=locs.modern$dist,
                                            pObs=locs.modern$throat,
                                            nEff=as.vector(table(hyb_Index.modern$Population)))

#run 3 models
throat.contemp<-run3hzarmodels(input.trait=throat.contemp.input,
                              begin=0,
                              end=1994)
```

```{r, fig.height=10, fig.width=6, fig.align = "center"}
#check convergence
check.convergence(throat.contemp)
```

```{r}
#run analysis function on genome wide ancestry input
contemp.plot<-analyze.hzar.output(throat.contemp, input.var = "contemporary throat color")
```

### plot clines overlaid
```{r}
#plot the clines overlaid
hzar.plot.cline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.cline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="gray")

#plot the clines overlaid with 95% credible intervals
hzar.plot.fzCline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.fzCline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="black",fzCol = "red")
```

### save clines overlaid
```{r}
pdf("magpiejay.overlaid.clines.pdf", width = 4.25, height = 4) #open PDF
#plot the clines overlaid
hzar.plot.cline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.cline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="gray")
dev.off() #close PDF

pdf("magpiejay.overlaid.fuzzy.clines.pdf", width = 4.25, height = 4) #open PDF
#plot the clines overlaid with 95% credible intervals
hzar.plot.fzCline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.fzCline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="gray")
dev.off() #close PDF
```

### plot 2 log-likelihood center estimates
```{r}
#get 2LL estimates of center for each selected model and add them to a dataframe for plotting
center.vals<-hzar.getLLCutParam(contemp.plot$analysis$model.selected, names(contemp.plot$analysis$model.selected$data.param))[1:2]
center.vals$center<-contemp.plot$analysis$modeldetails$param.all$center
center.vals$input<-c("contemporary")
center.vals[2,]<-c(hzar.getLLCutParam(hist.plot$analysis$model.selected, names(hist.plot$analysis$model.selected$data.param))[1:2],
                   hist.plot$analysis$modeldetails$param.all$center,
                   "historical")

#plot as box plots
boxplot(center.vals$center ~ center.vals$input, ylim = c(0, 2000), horizontal = TRUE)
rect(center.vals$center2LLLow[center.vals$input == "contemporary"],.8,
     center.vals$center2LLHigh[center.vals$input == "contemporary"],1.2, col="black")
rect(center.vals$center2LLLow[center.vals$input == "historical"],1.8,
     center.vals$center2LLHigh[center.vals$input == "historical"],2.2, col="gray")

#save box plots
pdf("magpiejay.gen.centers.pdf", width = 4.25, height = 3) #open PDF
boxplot(center.vals$center ~ center.vals$input, ylim = c(0, 2000), horizontal = TRUE)
rect(center.vals$center2LLLow[center.vals$input == "contemporary"],.8,
     center.vals$center2LLHigh[center.vals$input == "contemporary"],1.2, col="black")
rect(center.vals$center2LLLow[center.vals$input == "historical"],1.8,
     center.vals$center2LLHigh[center.vals$input == "historical"],2.2, col="gray")
dev.off()

#plot above the cline
#set plotting layout
par(mar = c(4, 4, .1, .1))
layout.matrix <- matrix(c(1, 2), nrow = 2, ncol = 1)

layout(mat = layout.matrix,
       heights = c(2, 5), # Heights of the two rows
       widths = c(8)) # Width of the column
#plot1
boxplot(center.vals$center ~ center.vals$input, ylim = c(0, 2000), horizontal = TRUE)
rect(center.vals$center2LLLow[center.vals$input == "contemporary"],.8,
     center.vals$center2LLHigh[center.vals$input == "contemporary"],1.2, col="black")
rect(center.vals$center2LLLow[center.vals$input == "historical"],1.8,
     center.vals$center2LLHigh[center.vals$input == "historical"],2.2, col="gray")
#plot2
#plot the clines overlaid
hzar.plot.cline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.cline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="gray")

#save plot
pdf("magpiejay.overlaid.genetic.clines.with.centers.pdf", width = 4.25, height = 5) #open PDF
#plot above the cline
#set plotting layout
par(mar = c(4, 4, .1, .1))
layout.matrix <- matrix(c(1, 2), nrow = 2, ncol = 1)

layout(mat = layout.matrix,
       heights = c(2, 5), # Heights of the two rows
       widths = c(8)) # Width of the column
#plot1
boxplot(center.vals$center ~ center.vals$input, ylim = c(0, 2000), horizontal = TRUE)
rect(center.vals$center2LLLow[center.vals$input == "contemporary"],.8,
     center.vals$center2LLHigh[center.vals$input == "contemporary"],1.2, col="black")
rect(center.vals$center2LLLow[center.vals$input == "historical"],1.8,
     center.vals$center2LLHigh[center.vals$input == "historical"],2.2, col="gray")
#plot2
#plot the clines overlaid
hzar.plot.cline(contemp.plot$analysis$model.selected,pch=24,xlab="Distance (km)")
hzar.plot.cline(hist.plot$analysis$model.selected,pch=24,xlab="Distance (km)",add=TRUE,col="gray")
dev.off() #save
```


#compare cline widths
```{r}
#pull point estimates and 2LL widths for modern and historical
center.vals<-hzar.getLLCutParam(contemp.plot$analysis$model.selected, names(contemp.plot$analysis$model.selected$data.param))[3:4]
center.vals$center<-contemp.plot$analysis$modeldetails$param.all$width
center.vals$input<-c("contemporary")
center.vals[2,]<-c(hzar.getLLCutParam(hist.plot$analysis$model.selected, names(hist.plot$analysis$model.selected$data.param))[3:4],
                   hist.plot$analysis$modeldetails$param.all$width,
                   "historical")

#show each width estimate and 2LL confidence band
center.vals

#make plotting df
df<-data.frame(val=c(t(center.vals)[1:3,1],t(center.vals)[1:3,2]),
               time=c(rep("cont", times=3),rep("hist",times=3)))
df$val<-as.numeric(df$val)
boxplot(val~time,data=df)

#plot
boxplot(center.vals$center ~ center.vals$input, horizontal = FALSE, ylim = c(0, 200), main="cline width comparison")
rect(.8,center.vals$width2LLLow[center.vals$input == "contemporary"],
     1.2,center.vals$width2LLHigh[center.vals$input == "contemporary"], col="black")
rect(1.8, center.vals$width2LLLow[center.vals$input == "historical"],
     2.2, center.vals$width2LLHigh[center.vals$input == "historical"], col="gray")
#I can't remember what the significance measure is here. Do the 2LL estimates need to be non-overlapping? Or is it good enough that the point estimates are reciprocally outside the 2LL estimate from the alternate time period?

#save
pdf("magpiejay.width.estimates.pdf", width = 5, height = 5) #open PDF
boxplot(center.vals$center ~ center.vals$input, horizontal = FALSE, ylim = c(0, 200), main="cline width comparison")
rect(.8,center.vals$width2LLLow[center.vals$input == "contemporary"],
     1.2,center.vals$width2LLHigh[center.vals$input == "contemporary"], col="black")
rect(1.8, center.vals$width2LLLow[center.vals$input == "historical"],
     2.2, center.vals$width2LLHigh[center.vals$input == "historical"], col="gray")
dev.off()
```

